Parallel Global Optimization for Non-Convex Mixed-Integer Problems \thanks{This study was supported by the Russian Science Foundation, project No.\,16-11-10150.}

### 2019-04-10 2019/RSD_Integer/RSD2019.tex
Before correction

\usepackage[utf8x]{inputenc}
\usepackage[english,russian]{babel}

@@

The paper text reflecting the results of the performed study is organized in the following way. 
In Section 2, a brief description of the dimensionality reduction scheme using the space-filling 
curves is given. Also, the index scheme of accounting for the constraints is described. Here the 
formulation of the parallel index algorithm for solving the continuous global optimization 
problems is given as well.
In Section 3, the approach to the generalization of the parallel index algorithm with the purpose 
of solving the mixed-integer problems is presented. A method allowing reducing the solving of 
mixed-integer problem to solving a set of the continuous optimization problems, which can be 
performed in parallel is given. 
Section 4 contains the results of numerical experiments. The comparison of the sequential 
version of the algorithm with the known analogs is conducted here. Also, the efficiency of the 
parallel CPU- and GPU-versions of the algorithm for solving a series of the multiextremal 
mixed-integer problem is demonstrated. 
Section 5 concludes the paper.

@@

An efficient parallel index algorithm (PIA) for solving the constrained global optimization 
problem (\ref{problem}) has been developed at University of Nizhnii Novgorod. The scheme of 
the algorithm is as follows.

@@

\section{Parallel algorithm for mixed-integer problems}
%Russian
Now let us consider the case when the argument of the problem functions consists of two 
components: the vector $y$ falling into the hyperinterval $D$ and the vector $u$ taking a finite 
(and not too large) set of possible values i. e. 
\begin{gather}\label{problem_i}
\min{\left\{ g_{m+1}(y,u):y\in D, \; g_i(y,u)\leq 0, \; 1 \leq i \leq m\right\}},\\
D=\left\{a_j\leq y_j \leq b_j, \; 1\leq j \leq N \right\}.\nonumber
\end{gather}

@@

Let us number by the integer values $s, 1\leq s \leq S,$ all possible values of the vector $u$ i. e. 
superimpose to each considered value $s$ to the vector $u_s$. 
Then, the considered problem can be written in the form 
\begin{gather}\label{problem_is}
 \min_{s\in\{1,...,S\}}\left(\min{\left\{ g_{m+1}(y,u_s):y\in D, \; g_i(y,u_s)\leq 0, \; 1 \leq i \leq 
m\right\}}\right),\\
D=\left\{ a_j\leq y_j \leq b_j, \; 1 \leq j\leq N \right\}.\nonumber 
\end{gather}

@@

Now let us consider the relation 
\[
Y(x)=y(x-E(x)), \; x\in[0,S],
\]
mapping any point of the interval [0,S] onto the domain $D$ (the notation $E(x)$ corresponds 
to the integer part of the number $x$) and define the functions 
\[
g_i(x) = g_i(Y(x),u_{E(x)+1}), x\in[0,S],
\]
having, in general, the discontinuities of the first kind at the integer points $x_k = i, 1\leq i \leq 
S-1$.
The values $z_k = g_\nu(y(x_k))$ from (\ref{trial_result}) at these points will be considered to 
be undefined, and the values of the indices -- to equal to 0 i. e. $\nu(x_k) = 0$.

@@

As an illustration, Fig. \ref{fig:1} presents the plots of the functions corresponding to a 
problem with one continuous parameter and one binary parameter.

@@

Thus, we have constructed the \textit{Mixed-Integer Parallel Index Algorithm} (MIPIA) based 
on the  
reduction of the mixed-integer non-convex optimization problem to the non-convex 
optimization problem. 
%Russian
The proposed computational scheme is based on the parallel index method for solving the 
continuous optimization problems and is not oriented onto any particular computational device. 
Here the parallel performing of several trials at different points of the search domain is one of 
key operations (see Step 7 of the algorithm), which can be implemented on CPU (with the use 
of OpenMP and/or MPI) as well as on GPU (with the use CUDA). The issues of the use of 
GPU for these purposes have been considered in details in 
Ref.~\cite{Barkalov2016,Gergel2016}. This approach to the use of GPU has been applied in 
the implementation of the MIPIA algorithm as well.

@@

The next series of experiments were performed in order to evaluate the speedup of the parallel 
version of the proposed MIPIA algorithm with the use of CPU as well as GPU. In these 
experiments, a series of 100 mixed-integer test problems generated in a random way was solved.

GKLS is a well known generator of the test problems for the continuous multiextremal 
optimization~\cite{Gaviano}. It allows generating the functions of arbitrary dimensionality with 
known properties (the number of local minima, the size of their domains of attraction, the global 
minimizer, etc.).  
This generator of multiextremal functions is used for the investigations of the global 
optimization algorithms 
often~\cite{Paulavicius2014,SergeyevKvasov2015,Lebedev2015,Gergel2015}. 
In Fig.~\ref{example} (a) and (b), the contour plots of two-dimensional GKLS functions are 
presented. Figures also shows the points of the trials performed by the method until the required 
accuracy $\epsilon=10^{-2}$ was achieved.

@@

In the performed experiments, the GKLS generator was used as a base for the construction the 
mixed-integer problems. The rules, allowing generating the test problems of this kind consist in 
the following.

\begin{enumerate}
	\item A continuous multiextremal function $\varphi(y), \; y\in D, \; D = \left\{ a_j\leq 
y_j\leq b_j, 1\leq j \leq N \right\}$ is generated with the use of the GKLS generator. The global 
minimum of this function is achieved at the known point $y'=(y'_1,...,y'_N)$ and equals 
$\varphi'=\varphi(y')=-1$.
	\item A concave mixed-integer function 
	\[
			\mu(y,u) = -2 \left[ \sum_{j=1}^N \left( \frac{y_j - y'_j}{b_j-a_j} 
\right)^2 + \sum_{j=1}^M \left( \frac{u_j - b_j}{b_j-a_j} \right)^2 \right],
	\]
	is generated where 
	\begin{gather}
	y\in D = \left\{ a_j\leq y_j\leq b_j, 1\leq j \leq N \right\} \subset R^N,\nonumber \\
	u\in U = \left\{ u_j \in  \left\{a_j, ..., b_j \right\}, 1\leq j \leq M \right\};\nonumber
	\end{gather}
	i. e. this function has $N$ continuous and $M$ discrete parameters and achieves its 
minimum value at the point $(y',b)$.
	\item The coefficient 
	\[
	C = 4 - \min_{y,u} \left\{ \mu(y,u) \right\}, \; y\in D, u \in U.
	\]
	is computed. 
It is obvious, that the minimum of a concave function $\mu(y,u)$ is located at one of the corner 
points of the search domain. Therefore, if the problem dimensionality is of the order of 10 it can 
be found by the brute force method.
	\item A multiextremal mixed-integer function 
	\[
	f(y,u) = \left(\varphi(y) + \sum_{j=1}^M{u_j}\right)\left(C + \mu(y,u)\right).
	\]
	is formed.
By construction, $f(y,u)$ would take its minimum value at the point $(y',b)$.

@@

Total 100 11-dimensional mixed-integer problems of this kind were generated. The accuracy of 
the search was equal to $10^{-2}$.   
Computational experiments were carried out on Lobachevsky supercomputer. The node of 
supercomputer included two Intel Sandy Bridge E5-2660 2.2 GHz CPUs and 64 Gb RAM. The 
CPU had 8 cores, i.e. each node had a total of 16 cores and two NVIDIA Kepler K20X GPUs.

In Table \ref{tab:2}, the results of experiments on CPU with the use of OpenMP are presented 
subject to the number of employed threads $p$. Total 1, 8, and 16 threads were used. The 
average number of iterations $K_{av}$ required to solve the problem, the average time of 
solving $T_{av}$ (in seconds), the time speedup $S$, and iteration speedup $s$ (with respect 
to the sequential run i. e. for $p=1$) are presented. In accordance with the parallelization 
scheme, the number of trials within a single iteration of the parallel algorithm was equal to the 
number of employed threads.

@@

As one can see from Table \ref{tab:2}, almost linear speedup in iterations and an essential 
speedup in the problem solving time have been observed. At that, the sequential algorithm spent 
almost 1.5 hours in average to solve a problem.

@@

\section{Conclusion}
%Russian
In the present paper, the results obtained in Lobachevskii State University of Nizhnii Novgorod 
when developing and investigating the parallel global optimization algorithms for solving the 
multiextremal problems, in which some parameters are continuous and the others are the discrete 
ones are presented. 
An efficient parallel algorithm has been proposed for solving the problems of the specified class. 
In the sequential variant, this algorithm does not inferior to similar purpose algorithm 
implemented in Matlab Global Optimization Toolbox.
In the parallel variant, the proposed algorithm allows the efficient implementations on CPU as 
well as on GPU. 
The numerical experiments
on a series of 100 multiextremal mixed-integer test problems have been carried out 
demonstrating a good speedup of the algorithm with the use of GPU convincingly.
Thus, the parallel algorithm with the use of GPU demonstrated the time speedup $S = 182$ 
relative to the sequential one and $S = 23.7$ relative to the algorithm employing two CPU on 
the cluster node fully.

>>>>

%\usepackage[utf8x]{inputenc}
%\usepackage[english,russian]{babel}

@@

The paper text reflecting the results of the performed study is organized in the following way. 
In Section 2, a brief description of the dimensionality reduction scheme using the space-filling 
curves is given. Also, the index scheme of accounting for the constraints is described. Here the 
formulation of the parallel index algorithm for solving the continuous global optimization 
problems is given as well.
In Section 3, the approach to the generalization of the parallel index algorithm with the purpose 
of solving the mixed-integer problems is presented. A method allowing reducing the solving of 
mixed-integer problem to solving a set of the continuous optimization problems, which can be 
performed in parallel, is given. 
Section 4 contains the results of numerical experiments. The comparison of the sequential 
version of the algorithm with the known analogs is conducted here. Also, the efficiency of the 
parallel CPU- and GPU-versions of the algorithm for solving a series of the multiextremal 
mixed-integer problem is demonstrated. 
Section 5 concludes the paper.

@@

An efficient parallel index algorithm (PIA) for solving the constrained global optimization 
problem (\ref{problem}) has been developed at University of Nizhni Novgorod. The scheme of 
the algorithm is as follows.

@@

\section{Parallel algorithm for mixed-integer problems}
%Russian
Now let us consider the case when the argument of the problem functions consists of two 
components: the vector $y$ from the hyperinterval $D$ and the vector $u$ taking a finite 
(and not too large) set of possible values, i.e. 
\begin{gather}\label{problem_i}
\min{\left\{ g_{m+1}(y,u):y\in D, \; g_i(y,u)\leq 0, \; 1 \leq i \leq m\right\}},\\
D=\left\{a_j\leq y_j \leq b_j, \; 1\leq j \leq N \right\}.\nonumber
\end{gather}

@@

Let us number by the integer values $s, 1\leq s \leq S,$ all possible values of the vector $u$, i.e. 
juxtapose each considered value $s$ with the vector $u_s$. 
Then, the considered problem can be written in the form 
\begin{gather}\label{problem_is}
 \min_{s\in\{1,...,S\}}\left(\min{\left\{ g_{m+1}(y,u_s):y\in D, \; g_i(y,u_s)\leq 0, \; 1 \leq i \leq 
m\right\}}\right),\\
D=\left\{ a_j\leq y_j \leq b_j, \; 1 \leq j\leq N \right\}.\nonumber 
\end{gather}

@@

Now let us consider the relation 
\[
Y(x)=y(x-E(x)), \; x\in[0,S],
\]
mapping any point of the interval [0,S] onto the domain $D$ (the notation $E(x)$ corresponds 
to the integer part of the number $x$) and define the functions 
\[
g_i(x) = g_i(Y(x),u_{E(x)+1}), x\in[0,S],
\]
having, in general, the jump discontinuities at the integer points $x_k = i, 1\leq i \leq 
S-1$.
The values $z_k = g_\nu(y(x_k))$ from (\ref{trial_result}) at these points will be considered to 
be undefined, and the values of the indices -- to equal to 0, i.e. $\nu(x_k) = 0$.

@@

As an illustration, Fig. \ref{fig:1} presents the plots of the functions corresponding to a 
problem 
\[
%\min{\left\{ x^2 - 10\cos(2\pi x) + u^2 - 10\cos(2\pi u) : x\in [-2.22, 1.8], u \in {0,1} \right\}}
\min{\left\{ u^2 (\sin(x) +\sin(10x/3) : x\in [2.7, 7.5], u \in {1,2} \right\}}
\]
with one continuous parameter $x$ and one binary parameter $u$.
%Еще один график?

@@

Thus, we have constructed the \textit{Mixed-Integer Parallel Index Algorithm} \break 
(MIPIA) based on the  reduction of the mixed-integer non-convex optimization problem to the non-convex optimization problem. 
%Russian
The proposed computational scheme is based on the parallel index method for solving the 
continuous optimization problems and is not oriented onto any particular computational device. 
Here the parallel performing of several trials at different points of the search domain is one of 
key operations (see Step 7 of the algorithm), which can be implemented on CPU (with the use 
of OpenMP and/or MPI) as well as on GPU (with the use CUDA). The issues of the use of 
GPU for these purposes have been considered in details in 
\cite{Barkalov2016,Gergel2016}. This approach to the use of GPU has been applied in 
the implementation of the MIPIA algorithm as well.

@@

The next series of experiments were performed in order to evaluate the speedup of the parallel 
version of the proposed MIPIA algorithm with the use of CPU as well as GPU. In these 
experiments, a series of 100 mixed-integer test problems generated in a random way was solved. 
Computational experiments were carried out on Lobachevsky supercomputer. The node of 
supercomputer included two Intel Sandy Bridge E5-2660 2.2 GHz CPUs and 64 Gb RAM. The 
CPU had 8 cores, i.e. each node had a total of 16 cores and two NVIDIA Kepler K20X GPUs.

GKLS~\cite{Gaviano} is a well known generator of the test problems for the continuous multiextremal 
optimization. It allows generating the functions of arbitrary dimensionality with 
known properties (the number of local minima, the size of their domains of attraction, the global 
minimizer, etc.).  
This generator of multiextremal functions is often used for the investigations of the global 
optimization algorithms ~\cite{Paulavicius2014,SergeyevKvasov2015,Lebedev2015,Gergel2015}. 
In Fig.~\ref{example} (a) and (b), the contour plots of two-dimensional GKLS functions are 
presented. Figures also shows the points of the trials performed by the method until the required 
accuracy $\epsilon=10^{-2}$ was achieved.

@@

In the performed experiments, the GKLS generator was used as a base for the construction the 
mixed-integer problems. The rules allowing generating the test problems of this type consist in 
the following.

\begin{enumerate}
	\item A continuous multiextremal function $\varphi(y), y\in D = \left\{ a_j\leq 
y_j\leq b_j, 1\leq j \leq N \right\}$ is generated with the use of the GKLS generator. The global 
minimum of this function is achieved at the known point $y'=(y'_1,...,y'_N)$ and equals 
$\varphi'=\varphi(y')=-1$.
	\item A concave mixed-integer function 
	\[
			\mu(y,u) = -2 \left[ \sum_{j=1}^N \left( \frac{y_j - y'_j}{b_j-a_j} 
\right)^2 + \sum_{j=1}^M \left( \frac{u_j - b_j}{b_j-a_j} \right)^2 \right],
	\]
	is generated where 
	\begin{gather}
	y\in D = \left\{ a_j\leq y_j\leq b_j, 1\leq j \leq N \right\} \subset R^N,\nonumber \\
	u\in U = \left\{ u_j \in  \left\{a_j, ..., b_j \right\}, 1\leq j \leq M \right\}.\nonumber
	\end{gather}
	This function has $N$ continuous and $M$ discrete parameters and achieves its 
minimum value at the point $(y',b)$.
	\item The coefficient 
	\[
	C = 4 - \min_{y,u} \left\{ \mu(y,u) \right\}, \; y\in D, u \in U.
	\]
	is computed. 
It is obvious, that the minimum of a concave function $\mu(y,u)$ is located at one of the corner 
points of the search domain. Therefore, if the problem dimensionality is of the order of 10 it can 
be found by the brute force method.
	\item A multiextremal mixed-integer function 
	\[
	f(y,u) = \left(\varphi(y) + \sum_{j=1}^M{u_j}\right)\left(C + \mu(y,u)\right).
	\]
	is formed.
By construction, $f(y,u)$ would take its minimum value at the point $(y',b)$.

@@

A hundred 11-dimensional mixed-integer problems of this type were generated in total. 
For the purpose of simulation of the computational complexity inherent to
applied problems of optimization, calculation of the objective function in all performed
experiments was made more complex by additional calculations without changing the type
of function and arrangement of its minima (series summation from 80 thousand elements).
The accuracy of the search was equal to $10^{-2}$.%Параметры метода!!!

In Table \ref{tab:2}, the results of experiments on CPU with the use of OpenMP are presented 
subject to the number of employed threads $p$. Total 1, 8, and 16 threads were used. The 
average number of iterations $K_{av}$ required to solve the problem, the average time of 
solving $T_{av}$ (in seconds), the time speedup $S$, and iteration speedup $s$ (with respect 
to the sequential run, i.e. for $p=1$) are presented. In accordance with the parallelization 
scheme, the number of trials within a single iteration of the parallel algorithm was equal to the 
number of employed threads.

@@

As one can see from Table \ref{tab:2}, almost linear speedup in iterations and an significant 
speedup in the problem solving time have been observed. At that, the sequential algorithm spent 
almost 1.5 hours in average to solve a problem.

@@

\section{Conclusion}
%Russian
In the present paper, the results obtained in Lobachevskii State University of Nizhni Novgorod 
when developing and investigating the parallel global optimization algorithms for solving the 
multiextremal problems, in which some parameters are continuous and the others are the discrete 
ones are presented. 
An efficient parallel algorithm has been proposed for solving the problems of the specified class. 
In the sequential variant, this algorithm does not inferior to similar purpose algorithm 
implemented in Matlab Global Optimization Toolbox.
In the parallel variant, the proposed algorithm allows the efficient implementations on CPU as 
well as on GPU. 
The numerical experiments on a series of 100 multiextremal mixed-integer test problems have been carried out 
convincingly demonstrating a good speedup of the algorithm with the use of GPU.
Thus, the parallel algorithm with the use of GPU demonstrated the time speedup $S = 182$ 
relative to the sequential one and $S = 23.7$ relative to the algorithm fully employing two CPU on 
the cluster node.
